---
title: "FAME_FINAL"
author: "Leah Cuthbertson"
date: "19/11/2020"
output: word_document
---

#Load packages and working directory
```{r, include=TRUE, echo=FALSE}
library(phyloseq)
library(ggplot2)
library(Biostrings) ##Note that Biostrings also has a distance function - load phyloseq second so its distance function is not masked. 
library(plyr)
library(stringr)
library(gridExtra)
library(reshape2)
library(knitr)
library(vegan)
library(decontam)
```

#Initial preprocessing bacterial data 

Data was obtained from processing of raw data through QIIME1.9

```{r, include=TRUE, echo=FALSE}
all_fame<-import_biom(
  "../data/seq_files/otu_table_hdf5.biom", "../data/seq_files/fame.tre", "../data/seq_files/rep_set_for_r.fasta"
)
#all_fame <- readRDS("../data/bacterial_phyloseq.rds")
data1<-read.csv("../data/fame_map.csv", row.names=1)
sample_data(all_fame)<-sample_data(data1)
sample_names(all_fame)<-sample_data(all_fame)$Description
```

##Preprocessing
Function from Aaron Saunders - does the relabelling
```{r}
makeTaxLabel <- function(OTU, mydata){
 # Makes a string label using the lowest informative tax level
 #
 # Args:
 #   OTU: OTU number
 #   mydata: the phyloseq object with the tax table
 #
 # Returns:
 #   a tax name
 OTU <- as.character(OTU)  # the OTU numbers are stored as character not integer!
 taxstrings <- as.character(tax_table(mydata)[OTU])
 empty_strings <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__")
 tax_name <- NA
 tax_level <- length(taxstrings)  # start at lowest tax level

 while(is.na(tax_name) | 
       (tax_name %in% empty_strings)){
   tax_name  <- taxstrings[tax_level]
   tax_level <- tax_level -1
 }
 tax_name
}

tax_table(all_fame) =gsub("s__uncultured_bacterium", as.character(NA), tax_table(all_fame))
tax_table(all_fame) =gsub("s__uncultured_organism", as.character(NA), tax_table(all_fame))
tax_table(all_fame) =gsub("g__uncultured", as.character(NA), tax_table(all_fame))

mynames = NULL
for (i in 1:length(taxa_names(all_fame))){
 mynames <- rbind(mynames, c(makeTaxLabel(taxa_names(all_fame)[i],all_fame)))
}

mynames = gsub("s__", "", mynames)
mynames = gsub("g__", "", mynames)
mynames = gsub("f__", "", mynames)
mynames = gsub("o__", "", mynames)
mynames = gsub("c__", "", mynames)
mynames = gsub("p__", "", mynames)

## Append the taxa names (ie, OTU ID) and add as a new column to the tax table of the original .biom

OTUID = str_c(mynames[,1],"_",seq(1,length(taxa_names(all_fame)),1))
tax_table(all_fame) <- cbind(tax_table(all_fame), mynames=OTUID)
head(tax_table(all_fame))

## Rename tax table headings to something real
colnames(tax_table(all_fame)) = c("Domain", "Phylum", "Class", "Order", "Family",  "Genus", "Species", "OTUID")
## Rename the taxa_names (frequently used as plot labels by phyloseq) to be the same as the new unique informative names
taxa_names(all_fame) <- OTUID
```

##Decontam
1. remove less than 20 reads
```{r}

df <- as.data.frame(sample_data(all_fame)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(all_fame)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize)) + geom_point()

taxa_sums(all_fame)

plot(sort(taxa_sums(all_fame)))

Ssum<-data.frame(taxa_sums(all_fame))
plot(Ssum[order(-Ssum$taxa_sums.all_fame.),])
#take a closer look at the low abundance OTUs
a1<-prune_taxa(taxa_sums(all_fame)<=100, all_fame)
Ssum1<-data.frame(taxa_sums(a1))
plot(Ssum1[order(-Ssum1$taxa_sums.a1.),])
#prune out the singletons
a2<-prune_taxa(taxa_sums(a1)<=50, a1)
Ssum1<-data.frame(taxa_sums(a2))
plot(Ssum1[order(-Ssum1$taxa_sums.a2.),])
# OTUs start to plateau at about 20 reads 
rm(Ssum)
rm(Ssum1)
rm(a1)
rm(a2)

all_fame = filter_taxa(all_fame, function(x) sum(x) > 20, TRUE) #remove OTUs with less than 20 read

```
2. Explore the contamination from Vibrio. Small number of reads outside the mock community. 
These will be removed later. 
```{r}
vibro = subset_taxa(all_fame, Genus=="g__Vibrio")
```
Sort data so Decontam works 
Missing qPCR data replaced with 1 
```{r}
data2<-data.frame(sample_data(all_fame))
data2$total_16S_extract[is.na(data2$total_16S_extract)] <- 0

data2$total_16S_extract[data2$total_16S_extract == 0] <- 1

sample_data(all_fame)<-data2

sample_data(all_fame)$total_16S_extract <- (get_variable(all_fame, "total_16S_extract"))
sample_data(all_fame)$Negative<-sample_data(all_fame)$Study == "XS"
```
Run Decontam package
```{r, include=FALSE}

library(decontam)
fame_samps<-prune_samples(sample_data(all_fame)$Study!="Negative" & sample_data(all_fame)$Study!="Mock", all_fame)

contamdf.freq <- isContaminant(fame_samps, method="frequency", conc="total_16S_extract")
head(contamdf.freq)

table(contamdf.freq$contaminant) #56 contaminants
which(contamdf.freq$contaminant)

plot_frequency(fame_samps, taxa_names(all_fame)[c(which(contamdf.freq$contaminant))], conc="total_16S_extract") + 
 xlab("DNA Concentration")
########################################
fame_neg<-prune_samples(sample_data(all_fame)$Study!="Mock", all_fame)
sample_data(fame_neg)$Negative
#ps.noncontam <- prune_taxa(!contamdf.freq$contaminant, fame_neg)


#Look at prevalence only
contamdf.prev <- isContaminant(fame_neg, method="prevalence", neg="Negative") 
table(contamdf.prev$contaminant)#30 contaminants
which(contamdf.prev$contaminant)

plot_frequency(fame_neg, taxa_names(fame_neg)[c(which(contamdf.prev$contaminant))], conc="total_16S_extract") + 
 xlab("DNA Concentration")
```
3.Remove the OTUs indicated by qPCR
```{r, include=FALSE}
all_fame <- prune_taxa(!contamdf.freq$contaminant, all_fame)
```

4.Remove known contamination
```{r}
all_fame = subset_taxa(all_fame, Domain != "Unassigned"|is.na(Domain))
all_fame = subset_taxa(all_fame, Order != "o__Rhodobacterales"|is.na(Order))
all_fame = subset_taxa(all_fame, Order != "o__Rhizobiales"|is.na(Order))
all_fame = subset_taxa(all_fame, Family != "f__Oxalobacteraceae"|is.na(Family))
all_fame = subset_taxa(all_fame, Phylum != "p__Cyanobacteria"|is.na(Phylum))
all_fame = subset_taxa( all_fame, Family != "f__mitochondria"|is.na(Family))
all_fame = subset_taxa(all_fame, Genus != "g__Vibrio"|is.na(Genus))
```
##Save working dataset
```{r, include=FALSE}
saveRDS(all_fame, file= "../data/bacterial_phyloseq.rds")
```


#Initial preprocessing fungal data
Add in the fungal data
```{r}
fung_all <- import_biom(BIOMfilename = "../data/seq_files/fungi_raw_data/fungi_otu_table_hdf5.biom", 
                     refseqfilename = "../data/seq_files/fungi_raw_data/rep_set_for_r.fasta")
 
data2<-read.csv("../data/all_FAME_metadata.csv", row.names=2)
sample_data(fung_all)=sample_data(data2)
sample_names(fung_all)=sample_data(fung_all)$ID.1

controls<-subset_samples(fung_all, sample_data(fung_all)$Group != "NAFD"&sample_data(fung_all)$Group != "NTM"&sample_data(fung_all)$Group != "FB"&sample_data(fung_all)$Group != "ABPA"&sample_data(fung_all)$Group != "IBX")
#Only the crosssectional data is included in this study. Subset the data so that the samples match the bacterial samples. 
fung_XS<-fung_all
sample_data(fung_XS)<-sample_data(all_fame)
fung_XS<-merge_phyloseq(fung_XS,controls)
```
##Preprocessing
Function from Aaron Saunders - does the relabelling
```{r}

mynames = NULL
for (i in 1:length(taxa_names(fung_XS))){
 mynames <- rbind(mynames, c(makeTaxLabel(taxa_names(fung_XS)[i],fung_XS)))
}

mynames = gsub("s__", "", mynames)
mynames = gsub("g__", "", mynames)
mynames = gsub("f__", "", mynames)
mynames = gsub("o__", "", mynames)
mynames = gsub("c__", "", mynames)
mynames = gsub("p__", "", mynames)

## Append the taxa names (ie, OTU ID) and add as a new column to the tax table of the original .biom

OTUID = str_c(mynames[,1],"_",seq(1,length(taxa_names(fung_XS)),1))
tax_table(fung_XS) <- cbind(tax_table(fung_XS), mynames=OTUID)

## Rename tax table headings to something real
colnames(tax_table(fung_XS)) = c("Domain", "Phylum", "Class", "Order", "Family",  "Genus", "Species", "OTUID")
## Rename the taxa_names (frequently used as plot labels by phyloseq) to be the same as the new unique informative names
taxa_names(fung_XS) <- OTUID

```

##Decontam
Sort data so Decontam works 
Missing qPCR data replaced with 1 so that it will work
```{r}
data2<-data.frame(sample_data(fung_XS))
data2$X18S_quantity_per_ul[is.na(data2$X18S_quantity_per_ul)] <- 0

data2$X18S_quantity_per_ul<-data2$X18S_quantity_per_ul+1


sample_data(fung_XS)<-sample_data(data2)
```
1. Run Decontam package
```{r, include=FALSE}
fung_samps<-prune_samples(sample_data(fung_XS)$Group!="Neg"&sample_data(fung_XS)$Group!="Mock"&sample_data(fung_XS)$Group!="kitneg", fung_XS)

contamdf.freq <- isContaminant(fung_samps, method="frequency", conc="X18S_quantity_per_ul")
head(contamdf.freq)

table(contamdf.freq$contaminant) #74 contaminants 
which(contamdf.freq$contaminant)

plot_frequency(fung_samps, taxa_names(fung_XS)[c(which(contamdf.freq$contaminant))], conc="X18S_quantity_per_ul") + 
 xlab("DNA Concentration")
########################################
sample_data(fung_XS)$Negative<-sample_data(fung_XS)$Group == "Neg"

fung_neg<-prune_samples(sample_data(fung_XS)$Group!="Positive"&sample_data(fung_XS)$Group!="kitcontrol", fung_XS)

ps.noncontam <- prune_taxa(!contamdf.freq$contaminant, fung_neg)

#Look at prevalence only
contamdf.prev <- isContaminant(fung_neg, method="prevalence", neg="Negative") #0 contaminants
table(contamdf.prev$contaminant)# 19 contaminants
which(contamdf.prev$contaminant)

plot_frequency(fung_neg, taxa_names(fung_neg)[c(which(contamdf.prev$contaminant))], conc="X18S_quantity_per_ul") + 
 xlab("DNA Concentration")
#########################################

contamdf.combo <- isContaminant(fung_XS, method="combined", conc="X18S_quantity_per_ul", neg = "Negative")
head(contamdf.combo)

table(contamdf.combo$contaminant) # 65 contaminants 
which(contamdf.combo$contaminant)

```
Remove the OTUs indicated by qPCR
```{r, include=FALSE}
fung_XS <- prune_taxa(!contamdf.freq$contaminant, fung_XS)
fung_XS
```

2. remove Batrachochytrium as only in the mocks

```{r}
fung_work<-subset_taxa(fung_XS,  Genus != "g__Batrachochytrium"| is.na(Genus))

```
3. remove less than 20 reads
```{r}

df <- as.data.frame(sample_data(fung_work)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(fung_work)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize)) + geom_point()


plot(sort(taxa_sums(fung_work)))

Ssum<-data.frame(taxa_sums(fung_work))
plot(Ssum[order(-Ssum$taxa_sums.fung_work.),])
#take a closer look at the low abundance OTUs
a1<-prune_taxa(taxa_sums(fung_work)<=100, fung_work)
Ssum1<-data.frame(taxa_sums(a1))
plot(Ssum1[order(-Ssum1$taxa_sums.a1.),])
#prune out the singletons
a2<-prune_taxa(taxa_sums(a1)<=50, a1)
Ssum1<-data.frame(taxa_sums(a2))
plot(Ssum1[order(-Ssum1$taxa_sums.a2.),])
# OTUs start to plateau at about 20 reads 
rm(Ssum)
rm(Ssum1)
rm(a1)
rm(a2)

fung_work = filter_taxa(fung_work, function(x) sum(x) > 20, TRUE) #remove OTUs with less than 20 read
```
Save working fungal data
```{r, echo=FALSE, include=FALSE}
saveRDS(fung_work, "../data/fungal_phyloseq.rds")
```