---
title: "FAME ecological anaylsis"
author: "Leah Cuthbertson"
date: "19/11/2020"
output: word_document
---

#Load packages and working directory
```{r, include=TRUE, echo=FALSE}
library(phyloseq)
library(ggplot2)
library(Biostrings) ##Note that Biostrings also has a distance function - load phyloseq second so its distance function is not masked. 
library(plyr)
library(stringr)
library(gridExtra)
library(reshape2)
library(knitr)
library(vegan)
library(decontam)
```

# BACTERIAL working data

Load working data

```{r, include=FALSE}
all_fame <- readRDS("../data/bacterial_phyloseq.rds")
```
controls

## Examine controls

1.Positive controls

```{r, include=FALSE}
pos <- prune_samples(sample_data(all_fame)$Study=="Mock", all_fame)
pos <- filter_taxa(pos, function(x) sum(x) != 0, TRUE)

pos_rare <- rarefy_even_depth(pos,min(sample_sums(pos)) )

pos_otu <- data.frame(t(otu_table(pos_rare)))
pos_data <- sample_data(pos_rare)
adonis(pos_otu~pos_data$Description, permutations = 999, method="bray")

estimate_richness(pos_rare)
```
2. negative controls

```{r, include=FALSE}
neg <- prune_samples(sample_data(all_fame)$Study=="Negative", all_fame)
neg <- filter_taxa(neg, function(x) sum(x) != 0, TRUE)
plot_bar(neg, fill="Genus")

```
#BACTERIAL DATA 

Remove the controls and samples with fewer than 2,000 reads

```{r}
fame_work <- prune_samples(sample_data(all_fame)$Study!="Negative"&sample_data(all_fame)$Study!="Mock", all_fame)
fame_work <- prune_samples(sample_data(fame_work)$Study=="XS", fame_work)
fame_work <- prune_samples(sample_sums(fame_work)>=2000, fame_work)
```
rarify bacterial data

```{r}
fame_work1 <- filter_taxa(fame_work, function(x) sum(x) != 0, TRUE)
fame_work1 <- prune_samples(sample_sums(fame_work1)>=2000, fame_work1)
fame_rare <- rarefy_even_depth(fame_work, min(sample_sums(fame_work1)), rngseed = 382)
sample_data(fame_rare)$Group[sample_data(fame_rare)$Group=="IBX"] <- "NAFD"
```
<!-- Save data -->
<!-- ```{r, echo=FALSE, include=FALSE} -->
<!-- save(fame_rare, file= "FAME_rare.Rdata") -->
<!-- ``` -->

#FUNGAL working data

Load fugnal working data

```{r, include=FALSE}
fung_work <- readRDS("../data/fungal_phyloseq.rds")
fung_work<-prune_samples(sample_data(fung_work)$Study=="XS", fung_work)
fung_work <- prune_samples(sample_sums(fung_work)>=2000, fung_work)
```
rarify fungal data
```{r}
fung_work1 <- filter_taxa(fung_work, function(x) sum(x) != 0, TRUE)
fung_rare <- rarefy_even_depth(fung_work1, min(sample_sums(fung_work1)), rngseed = 2384)
fung_rare <- prune_samples(sample_data(fung_rare)$Disease!="", fung_rare)
```

<!-- Save data -->
<!-- ```{r, echo=FALSE, include=FALSE} -->
<!-- save(fung_rare, file= "Fung_rare.Rdata") -->
<!-- ``` -->

# Ecological analysis

Note: the following samples are excluded:
* FAME0008 - excluded due to missing data
* FAME0038 - missing culture data

```{r}
sample_data(fame_rare)$Group[sample_data(fame_rare)$Group=="IBX"] <- "NAFD"

#remove NTM patient with missing culture data
fame_rare <- subset_samples(fame_rare, sample_names(fame_rare) != "FAME00008")

sample_data(fung_rare)$Group[sample_data(fung_rare)$Group=="IBX"] <- "NAFD"

#remove NTM patient with missing NTM culture data
fung_rare <- subset_samples(fung_rare, sample_names(fung_rare) != "FAME00008"&sample_names(fung_rare) != "FAME00381")

sample_data(fame_rare)$Gender[sample_data(fame_rare)$Gender=="0"] <- "f"
sample_data(fung_rare)$Gender[sample_data(fung_rare)$Gender=="0"] <- "f"

```

Summary stats of the two OTU tables

```{r}
summary(sample_sums(fung_work))
sd(sample_sums(fung_work)) /sqrt(length(sample_sums(fung_work)))

summary(sample_sums(fame_work))
sd(sample_sums(fame_work)) /sqrt(length(sample_sums(fame_work)))
```


# Bacterial and fungal diversity

## 1. CF vs Bx

### Bacterial diversity

```{r}
sample_data(fame_rare) <- cbind(sample_data(fame_rare), estimate_richness(fame_rare, measures = c("Observed", "Shannon", "Simpson")))
fame_data <- data.frame(sample_data(fame_rare))
fame_data[fame_data == ""] <- NA

fame_data$X16S_quantity_mean<-as.numeric(as.character(fame_data$X16S_quantity_mean))
fame_data$log_biomass<-log10(fame_data$X16S_quantity_mean)

# CF vs Bx 
wilcox.test(Observed~Disease, fame_data)
wilcox.test(Shannon~Disease, fame_data)
wilcox.test(Simpson~Disease, fame_data)
wilcox.test(X16S_quantity_mean~Disease, fame_data)
```
Fungal diversity

```{r}
sample_data(fung_rare) <- cbind(sample_data(fung_rare), estimate_richness(fung_rare, measures = c("Observed", "Shannon", "Simpson")))

fung_data <- data.frame(sample_data(fung_rare))
fung_data[fung_data == ""] <- NA

# CF vs Bx 
wilcox.test(Observed~Disease, fung_data)
wilcox.test(Shannon~Disease, fung_data)
wilcox.test(Simpson~Disease, fung_data)
wilcox.test(X18S_quantity_per_ul~Disease, fung_data)
``` 

## beta diversity between groups
```{r}
fame_otu<- data.frame(t(otu_table(fame_rare)))
fame_data <- sample_data(fame_rare)
adonis(fame_otu~fame_data$Disease, permutations = 999, method="bray")

adonis(fame_otu~fame_data$Disease, strata = fame_data$Group, permutations = 999, method="bray")

### add in the treatment varibales as a strata
fung_otu <- data.frame(t(otu_table(fung_rare)))
fung_data<- sample_data(fung_rare)


adonis(fung_otu~fung_data$Disease, permutations = 999, method="bray")
adonis(fung_otu~fung_data$Disease, strata = fung_data$Group, permutations = 999, method="bray")

```

# 2.Fungal disease groups 

Bacteria

```{r}
#Group
library(FSA)
library(DescTools)
library(dunn.test)

fame_data <- data.frame(fame_data)
fame_data$Group <- as.factor(fame_data$Group)

kruskal.test(Observed~Group, fame_data)
DunnTest(Observed~Group, fame_data, method="hochberg")

kruskal.test(Shannon~Group, fame_data)
DunnTest(Shannon~Group, fame_data, method="hochberg")

kruskal.test(Simpson~Group, fame_data)
DunnTest(Simpson~Group, fame_data, method="hochberg")


```

Fungal

```{r}
# Column "Group" in metadata
fung_data <- data.frame(fung_data)
fung_data$Group <- as.factor(fung_data$Group)

kruskal.test(Observed~Group, fung_data)
DunnTest(Observed~Group, fung_data,method="hochberg")

kruskal.test(Shannon~Group, fung_data)
DunnTest(Shannon~Group, fung_data,method="hochberg")

kruskal.test(Simpson~Group, fung_data)
DunnTest(Simpson~Group, fung_data,method="hochberg")
```

Merge fungal diversity to fame_data

```{r}
fung_div <- cbind(sample_names(fung_rare), estimate_richness(fung_rare, measures = c("Observed", "Shannon", "Simpson")))

names(fung_div) <- c("ID", "fung_Observed", "fung_Shannon", "fung_Simpson")

combined_data<-merge(fame_data, fung_div, by.x = "Description", by.y = "ID")

row.names(combined_data)<-combined_data$Description
sample_data(fame_rare)<- sample_data(combined_data)
```

diversity tests between disease groups - bacteria

```{r}
sample_data(fame_rare)$disease_groups <- paste(sample_data(fame_rare)$Disease,"_",sample_data(fame_rare)$Group)
sample_data(fung_rare)$disease_groups <- paste(sample_data(fung_rare)$Disease,"_",sample_data(fung_rare)$Group)

fame_data <- data.frame(sample_data(fame_rare))
fung_data <- data.frame(sample_data(fung_rare))
fame_data$Group <- as.factor(fame_data$Group)

kruskal.test(Observed~Group, fame_data)
DunnTest(Observed~Group, fame_data, method="hochberg")

kruskal.test(Shannon~Group, fame_data)
DunnTest(Shannon~Group, fame_data, method="hochberg")

kruskal.test(Simpson~Group, fame_data)
DunnTest(Simpson~Group, fame_data, method="hochberg")

fame_data$disease_groups <- as.factor(fame_data$disease_groups)
kruskal.test(Observed~disease_groups, fame_data)
DunnTest(Observed~disease_groups, fame_data,method="hochberg")

kruskal.test(Shannon~disease_groups, fame_data)
DunnTest(Shannon~disease_groups, fame_data,method="hochberg")

kruskal.test(Simpson~disease_groups, fame_data)
DunnTest(Simpson~disease_groups, fame_data,method="hochberg")
```

diversity tests between disease groups - fungi


```{r}
fung_data$Group <- as.factor(fung_data$Group)
kruskal.test(Observed~Group, fung_data)
DunnTest(Observed~Group, fung_data,method="hochberg")

kruskal.test(Shannon~Group, fung_data)
DunnTest(Shannon~Group, fung_data,method="hochberg")

kruskal.test(Simpson~Group, fung_data)
DunnTest(Simpson~Group, fung_data,method="hochberg")

fung_data$disease_groups <- as.factor(fung_data$disease_groups)
kruskal.test(Observed~disease_groups, fung_data)
DunnTest(Observed~disease_groups, fung_data,method="hochberg")

kruskal.test(Shannon~disease_groups, fung_data)
DunnTest(Shannon~disease_groups, fung_data,method="hochberg")

kruskal.test(Simpson~disease_groups, fung_data)
DunnTest(Simpson~disease_groups, fung_data,method="hochberg")
```
just CF group
```{r}
fame_cf <- subset_samples(fame_rare, Disease == "CF")
fung_cf <- subset_samples(fung_rare, Disease == "CF")

famecf_data <- data.frame(sample_data(fame_cf))
fungcf_data <- data.frame(sample_data(fung_cf))

data2 <- data.frame(sample_data(all_fame))
data2$total_16S_extract[is.na(data2$total_16S_extract)] <- 0
data2$total_16S_extract[data2$total_16S_extract == 0] <- 1

sample_data(all_fame) <-data2

famecf_data$Group <- as.factor(famecf_data$Group)
kruskal.test(total_16S_extract~Group, famecf_data)
DunnTest(total_16S_extract~Group, famecf_data, method="hochberg")

kruskal.test(Observed~Group, famecf_data)
DunnTest(Observed~Group, famecf_data, method="hochberg")

kruskal.test(Shannon~Group, famecf_data)
DunnTest(Shannon~Group, famecf_data, method="hochberg")

kruskal.test(Simpson~Group, famecf_data)
DunnTest(Simpson~Group, famecf_data, method="hochberg")


fungcf_data$Group <- as.factor(fungcf_data$Group)
fungcf_data$X18S_quantity_per_ul<-fungcf_data$X18S_quantity_per_ul

kruskal.test(X18S_quantity_per_ul~Group, fungcf_data)
DunnTest(X18S_quantity_per_ul~Group, fungcf_data,method="hochberg")

kruskal.test(Observed~Group, fungcf_data)
DunnTest(Observed~Group, fungcf_data,method="hochberg")

kruskal.test(Shannon~Group, fungcf_data)
DunnTest(Shannon~Group, fungcf_data,method="hochberg")

kruskal.test(Simpson~Group, fungcf_data)
DunnTest(Simpson~Group, fungcf_data,method="hochberg")
```

adonis by group
```{r}
fame_cf <- subset_samples(fame_rare, Disease == "CF")
fung_cf <- subset_samples(fung_rare, Disease == "CF")

famecf_data <- data.frame(sample_data(fame_cf))
fungcf_data <- data.frame(sample_data(fung_cf))

famecf_otu <- t(otu_table(fame_cf))
fungcf_otu <- t(otu_table(fung_cf))


adonis(famecf_otu~famecf_data$Group, permutations = 999, method="bray")
adonis(fungcf_otu~fungcf_data$Group, permutations = 999, method="bray")

```

#3. CFPE

alpha and beta diversity tests

```{r}

famecf_data$Exacing <- as.factor(famecf_data$Exacing)
wilcox.test(Observed~Exacing, famecf_data)
wilcox.test(Shannon~Exacing, famecf_data)
wilcox.test(Simpson~Exacing, famecf_data)
wilcox.test(X16S_quantity_mean~Exacing, famecf_data)


wilcox.test(Observed~Exacing, fungcf_data)
wilcox.test(Shannon~Exacing, fungcf_data)
wilcox.test(Simpson~Exacing, fungcf_data)
wilcox.test(X16S_quantity_mean~Exacing, fungcf_data)

adonis(fungcf_otu~fungcf_data$Exacing, permutations = 99999, method="bray")
```

